---
title: "Open-Air Baseball Parks"
subtitle: "The Impacts of Weather Conditions"
author: "Jason M Graham"
institute: "University of Scranton"
format: revealjs
bibliography: presentation_bib.bib
---

```{r}
#| echo: false
#| include: false

# load packages
library(tidyverse) # wrangling, plots etc.
library(GGally) # paired plots
library(skimr) # data summary
library(knitr) # for table formatting
library(kableExtra) # for extra table formatting 
library(patchwork) # for combining plots
library(RColorBrewer) # for plot colors

# set plot theme
theme_set(theme_minimal(base_size = 12))

venues_html <- "https://raw.githubusercontent.com/jmgraham30/baseball_weather/main/data/USProSportsVenues.csv"

bos_game_weather_html <- "https://raw.githubusercontent.com/jmgraham30/baseball_weather/main/data/boston_game_weather.csv"

# load map data for US states minus AK and HI
states_df <- map_data("state") %>%
  filter(region != "alaska" | region != "hawaii")

# load weather and game data for Boston 
boston_game_weather <- read_csv(bos_game_weather_html)
# create separate columns with day, month, and year corresponding
# to game dates; and create a binary win (1) or not (0) column
boston_game_weather <- boston_game_weather %>%
  mutate(year=year(date),
         month=month(date),
         day=day(date),
         wl_binary=ifelse(wl=="W",1,0))
# load sports venue data
pro_sports_venues <- read_csv(venues_html)
# extract baseball parks
baseball_parks <- pro_sports_venues %>%
  filter(Sport == "MLB")

# add info about open roof
baseball_parks$Open_Roof <- c("No",rep("Yes",9),
                              "No","Yes","Yes","Yes",
                              "No","No",
                              rep("Yes",7),"No","Yes",
                              "Yes","No","No","No","Yes")
```

## Initial Research Question

::: {.incremental}

- Each of the thirty Major League Baseball (MLB) teams plays in one of thirty stadiums or baseball parks.

- Of the thirty MLB baseball stadiums, 22 of them are open-air.

- **We would like to know if changing climate and weather patterns might have an effect on either team performance or game attendance.**

:::


## MLB Parks 

```{r}

ggplot() +
  geom_polygon(data=states_df,
               aes(long,lat,group=group),
               fill="white",color="darkgrey") + 
  geom_point(data=baseball_parks,aes(x=Longitude,y=Latitude,color=Open_Roof),
             size=2.5) +
  geom_point(data=baseball_parks %>% filter(Abbreviation == "BOS"),
                                            aes(x=Longitude,y=Latitude),size=5,color="darkblue") + 
  geom_label( 
    data=baseball_parks %>% filter(Abbreviation == "BOS"), # Filter data first
    aes(label="Fenway",x=-71,y=43.6),color = "black",
    fill="red") + 
  scale_color_manual(values=c("orange","darkblue")) + 
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank(),
        plot.title = element_text(hjust = 0.5)) + 
  xlab("") + ylab("") + ggtitle("Major League Baseball Parks")
```


## Fenway Park an Open-Air Park

![Image from [@ticketmasterStepInside]](https://blog.ticketmaster.com/wp-content/uploads/Fenway-Park-field.jpg){fig-align="center"}


## Fenway Home Game and Weather Data

```{r}
boston_game_weather %>%
  head(3) %>%
  kable()
```


## Game Outcome Variables

```{r}
#| message: false


ra_1 <- boston_game_weather %>%
  ggplot(aes(x=hm_runs)) + 
  geom_histogram(color="black",fill="lightblue") + 
  xlab("Number of Runs by Boston") + 
  ylab("Count")
# Number of visitor runs
ra_2 <- boston_game_weather %>%
  ggplot(aes(x=vis_runs)) + 
  geom_histogram(color="black",fill="lightblue") + 
  xlab("Number of Runs by Visitors") + 
  ylab("Count")
# Boston runs - Visitor runs
ra_3 <- boston_game_weather %>%
  ggplot(aes(x=score_diff,fill=wl)) + 
  geom_histogram(color="black") + 
  scale_fill_manual(values = c("#E69F00","#009E73")) + 
  labs(x = "Boston Runs - Visitor Runs",
       y = "Count",
       fill="Win/Not")
# Win or Not Win
ra_4 <- boston_game_weather %>%
  ggplot(aes(x=wl,fill=wl)) + 
  geom_bar(color="black") + 
  scale_fill_manual(values = c("#E69F00","#009E73")) + 
  theme(legend.position = "none") +
  labs(x = "No Win or Win",
       y = "Count")

(ra_1 + ra_2) / (ra_4 + ra_3)
```

## Game Attendance and Duration

```{r}
#| message: false

rb_1 <- boston_game_weather %>%
  ggplot(aes(x=duration)) + 
  geom_histogram(color="black",fill="lightblue") + 
  labs(x = "Game Duration",
       y = "Count")
# Game attendances
rb_2 <- boston_game_weather %>%
  ggplot(aes(x=attendance)) + 
  geom_histogram(color="black",fill="lightblue") + 
  labs(x = "Game Attendance",
       y = "Count")

(rb_1 + rb_2)
```




## References
